**Selective DNA and Protein Isolation from Marine Macrophyte Surfaces**

### 2020-11-16 submission/manuscript.Rmd
Coauthors review done.

\vspace{20mm}
Marino Korlević^1$*$^, Marsej Markovski^1^, Zihao Zhao^2^, Gerhard J. Herndl^2,3^, Mirjana Najdek^1^

@@

@@

\newpage
\linenumbers
\sisetup{mode=text}
\setlength\parindent{24pt}

## Summary
Studies of unculturable microbes often combine methods based on DNA, such as 16S rRNA sequencing and metagenomics, with methods that allow insight into the metabolic status, such as metaproteomics. To apply these techniques to the microbial community inhabiting the surfaces of marine macrophytes it is advisable to perform, prior to the analysis, a selective DNA and protein isolation so that the host material, present in higher quantities, is not hampering the analysis. Two protocols, for DNA and protein isolation, were adapted for selective extractions of DNA and proteins from epiphytic communities inhabiting the surfaces of two marine macrophytes: the seagrass *Cymodocea nodosa* and the macroalga *Caulerpa cylindracea*. Protocols showed an almost complete removal of the epiphytic community regardless of the sampling season, station, settlement or host species. The obtained DNA was suitable for metagenomic and 16S rRNA sequencing, while isolated proteins could be identified by mass spectrometry, showing that protocols can be used in 16S rRNA, metagenomic and metaproteomic analysis. Low presence of host DNA and proteins, observed in isolated samples, indicated a selective nature of the protocols. Furthermore, the procedures are based on universally available laboratory chemicals making the protocols widely applicable. Taken together, the adapted protocols ensure an almost complete removal of the macrophyte epiphytic community, are selective for microbes inhabiting macrophyte surfaces and are providing DNA and proteins applicable in 16S rRNA sequencing, metagenomics and metaproteomics.

\newpage
## Introduction
Surfaces of marine macrophytes are inhabited by a diverse microbial community whose structure and function are poorly understood [@Egan2013]. As less than 1 % of all prokaryotic species are culturable, to study these organisms, molecular methods such as 16S rRNA sequencing, metagenomics and metaproteomics are indispensable [@Amann1995; @Su2012]. Applying these techniques requires an initial isolation step, with the purpose of obtaining high quality DNA and proteins.

Biological material (i.e. proteins and DNA) from pelagic microbial communities is usually isolated by collecting cells onto filters and subsequently isolating the aimed compound [@Gilbert2009]. If a specific microbial size fraction is aimed sequential filtration is applied [@Massana1997; @Andersson2010]. In contrast, obtaining biological materials from microorganism inhabiting surfaces requires either a cell detachment procedure prior to isolation or the host material is coextracted together with the targeted material. Methods for separating microbial cells form the host include shaking of host tissue [@Gross2003; @Noges2010], scraping of macrophyte surfaces [@Uku2007] or the application of ultrasonication [@Weidner1996; @Cai2014]. It was shown that shaking alone is not sufficient to remove microbial cells, at least from plant root surfaces [@Richter-Heitmann2016]. Manual separation methods, such as scraping and brushing, are time consuming and subjective, as the detachment efficiency depends on host tissue and the person performing the procedure [@Cai2014]. Ultrasonication was proposed as an alternative method as it is providing better results in terms of detachment efficiency [@Cai2014; @Richter-Heitmann2016]. The downside of this procedure is that complete cell removal was still not obtained and tissue disruption was observed especially after the application of probe ultrasonication [@Richter-Heitmann2016]. An alternative to these cell detachment procedures is the isolation of targeted epiphytic compounds together with host materials [@Staufenberger2008; @Jiang2015]. This procedure can lead to problems in the following processing steps such as mitochondrial and chloroplast 16S rRNA sequence contaminations from the host [@Longford2007; @Staufenberger2008]. In addition, when performing metagenomics and metaproteomics host material can cause biased results towards more abundant host DNA and proteins.

An alternative to these procedures is a direct isolation of the targeted material by incubating macrophyte tissues in an extraction buffer. After the incubation, the undisrupted host tissue is removed and the isolation procedure continues, omitting host material contaminations. To our knowledge the only procedure describing a direct and selective epiphytic DNA isolation from the surfaces of marine macrophytes was provided by @Burke2009. In contrast to previously described methods this protocol enables an almost complete removal of the surface community and it was used for 16S rRNA gene clone library construction [@Burke2011a] and metagenome sequencing [@Burke2011]. This method, though providing a selective isolation procedure, uses a rapid multienzyme cleaner (3M) that is not available worldwide and without a known composition [@Burke2009]. Also to our knowledge, no selective isolation protocol for proteins from epiphytic communities inhabiting marine macrophytes was established.

In the present study, we adapted a protocol widely used for DNA isolation from filters [@Massana1997] and a protocol used for protein isolation from sediments [@Chourey2010; @Hultman2015] for selective extractions of DNA and proteins from epiphytic communities inhabiting the surfaces of two marine macrophytes: the seagrass *Cymodocea nodosa* and the macroalga *Caulerpa cylindracea*. In addition, we tested the removal efficiency of the protocol and the suitability of obtained DNA and proteins for 16S rRNA sequencing, metagenomics and metaproteomics.

@@

To assess the removal efficiency of DNA and protein isolation procedures leaves and thalli were examined under a confocal microscope before and after treatments were performed. Developed procedures showed an almost complete removal of the surface community of both, *C. nodosa* and *C. cylindracea*. In addition, a similar removal efficiency was observed for communities sampled in contrasting seasons, December 2017 (\autoref{confocal_december}) and June 2018 (\autoref{confocal_june}). Also, no effect of station, settlement or isolation procedure (DNA or protein) on the removal efficiency was observed (Figs. \ref{confocal_december} and \ref{confocal_june}).

To evaluate if the obtained DNA is suitable to determine the microbial community structure an Illumina sequencing of the V4 16S rRNA region was performed. Sequencing yielded a total of `r number_sequences` sequences after quality curation and exclusion of eukaryotic, mitochondrial and no relative sequences. The number of sequences classified as chloroplast was `r number_chloroplast_sequences`. After excluding these sequences the total number of retrieved reads dropped to `r number_sequences_without_chloroplast`, ranging from `r min_max$min` to `r min_max$max` sequences per sample (\autoref{supp-nseq_notus}). Even when the highest sequencing effort was applied the rarefaction curves did not level off which is commonly observed in high-throughput 16S rRNA amplicon sequencing procedures (\autoref{supp-rarefaction}). Sequence clustering at a similarity level of 97 \si{\percent} yielded a total of `r no_otus` different OTUs. Taxonomic classification of reads allowed for a macrophyte associated epiphytic community determination that was mainly composed of: *Alphaproteobacteria* (`r filter(plot, taxon=="Alphaproteobacteria")$mean` ± `r filter(plot, taxon=="Alphaproteobacteria")$sd` \si{\percent}), *Bacteroidota* (`r filter(plot, taxon=="Bacteroidota")$mean` ± `r filter(plot, taxon=="Bacteroidota")$sd` \si{\percent}), *Gammaproteobacteria* (`r filter(plot, taxon=="Gammaproteobacteria")$mean` ± `r filter(plot, taxon=="Gammaproteobacteria")$sd` \si{\percent}), *Desulfobacterota* (`r filter(plot, taxon=="Desulfobacterota")$mean` ± `r filter(plot, taxon=="Desulfobacterota")$sd` \si{\percent}), *Cyanobacteria* (`r filter(plot, taxon=="Cyanobacteria")$mean` ± `r filter(plot, taxon=="Cyanobacteria")$sd` \si{\percent}) and *Planctomycetota* (`r filter(plot, taxon=="Planctomycetota")$mean` ± `r filter(plot, taxon=="Planctomycetota")$sd` \si{\percent}) (\autoref{community}).

Primers used in this study, as in many other 16S rRNA amplicon procedures, also amplified chloroplast 16S rRNA genes. We observed a high proportion of chloroplast sequences in all analyzed samples (`r filter(plot, taxon=="Chloroplast")$mean` ± `r filter(plot, taxon=="Chloroplast")$sd` \si{\percent}) (\autoref{community}). To determine if chloroplast sequences originate from hosts or eukaryotic epiphytic organisms, we exported SILVA classified chloroplast sequences and reclassified them using the RDP (Ribosomal Database Project) training set that allows for a more detailed chloroplast classification. The largest proportion of sequences were classified as Bacillariophyta (`r filter(chloroplast, taxon=="Bacillariophyta")$mean` ± `r filter(chloroplast, taxon=="Bacillariophyta")$sd` \si{\percent}) indicating that the DNA removal procedure did not coextract larger quantities of host DNA (\autoref{chloroplast}). Chloroplast sequences classified as Streptophyta constituted `r streptophyta_16S$mean` ± `r streptophyta_16S$sd` \si{\percent} of all chloroplast sequences originating from *C. nodosa* samples, while sequences classified as Chlorophyta comprised only `r chlorophyta_16S$mean` ± `r chlorophyta_16S$sd` \si{\percent} of all chloroplast sequences associated with *C. cylindracea* samples.

To determine if the extracted DNA can be used for metagenomic sequencing four samples containing epiphytic DNA were selected and shotgun sequenced using an Illumina platform. Metagenomic sequencing yielded from `r format(min(metadata_metagenomic_statistics$raw_R1), big.mark=',')` to `r format(max(metadata_metagenomic_statistics$raw_R1), big.mark=',')` sequence pairs (\autoref{supp-metagenomic_statistics}). Obtained sequences were successfully assembled into contigs whose L50 ranged from `r format(min(metadata_metagenomic_statistics$L50), big.mark=',')` to `r format(max(metadata_metagenomic_statistics$L50), big.mark=',')` bp. In addition, predicted coding sequences were successfully functionally annotated (`r format(min(metadata_metagenomic_statistics$eggnog), big.mark=',')` -- `r format(max(metadata_metagenomic_statistics$eggnog), big.mark=',')` annotated sequences; \autoref{cog}) and taxonomically classified. Functional annotation allowed for an assessment of the relative contribution of each COG (Clusters of Orthologous Groups) functional category to the total number of annotated coding sequences (\autoref{cog}a). Functional categories containing the highest number of sequences were: `r cog_relative$"COG Functional Category"[3]` (`r cog_relative$Name[3]`), `r cog_relative$"COG Functional Category"[4]` (`r cog_relative$Name[4]`), `r cog_relative$"COG Functional Category"[5]` (`r cog_relative$Name[5]`), `r cog_relative$"COG Functional Category"[6]` (`r cog_relative$Name[6]`) and `r cog_relative$"COG Functional Category"[7]` (`r cog_relative$Name[7]`). If host DNA was coextracted with epiphytic it should be detected in larger proportions in sequenced metagenomes. Indeed, no higher proportions of coding sequences classified into phylum Streptophyta or Chlorophyta were detected (\autoref{supp-metagenomic_taxonomy}). Sequenced metagenomic DNA originating from the surface of *C. nodosa* contained `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCyM" & phylum=="Streptophyta")$cds_relative` \si{\percent} of coding sequences classified into the phylum Streptophyta in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCyM" & phylum=="Streptophyta")$cds_relative` \si{\percent} in June 2018. Furthermore, the summed RPKM (Reads Per Kilobase Million) of these sequences constituted `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCyM" & phylum=="Streptophyta")$rpkm_relative` \si{\percent} of total RPKM of all successfully classified sequences in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCyM" & phylum=="Streptophyta")$rpkm_relative` \si{\percent} in June 2018. Similar low proportions of host coding sequences were detected in metagenomic samples originating from the surfaces of *C. cylindracea*. Of all successfully classified coding sequences `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCa" & phylum=="Chlorophyta")$cds_relative` \si{\percent} were classified into Chlorophyta in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCa" & phylum=="Chlorophyta")$cds_relative` \si{\percent} in June 2018. A relatively higher proportion of these sequences' RPKM than in the case of *C. nodosa* was observed, indicating a higher coextraction of host DNA in *C. cylindracea*. In December, the proportion of RPKM of sequences classified into Chlorophyta was `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCa" & phylum=="Chlorophyta")$rpkm_relative` \si{\percent}, while in June 2018 it reached `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCa" & phylum=="Chlorophyta")$rpkm_relative` \si{\percent}.

To evaluate if the procedure for protein extraction is suitable for metaproteomic analysis, obtained proteins were trypsin digested and sequenced using a mass spectrometer. Obtained MS/MS spectra were searched against a protein database from sequenced metagenomes. From `r format(min(metap_cog_num), big.mark=',')` to `r format(max(metap_cog_num), big.mark=',')` proteins were identified in isolated protein samples (\autoref{cog}b). In addition, successful identification of proteins allowed for an assessment of the relative contribution of each COG functional category to the total number of identified proteins (\autoref{cog}b). Functional categories containing the highest number of identified proteins were: `r metap_cog_relative$"COG Functional Category"[1]` (`r metap_cog_relative$Name[1]`), `r metap_cog_relative$"COG Functional Category"[2]` (`r metap_cog_relative$Name[2]`), `r metap_cog_relative$"COG Functional Category"[3]` (`r metap_cog_relative$Name[3]`), `r metap_cog_relative$"COG Functional Category"[4]` (`r metap_cog_relative$Name[4]`) and `r metap_cog_relative$"COG Functional Category"[5]` (`r metap_cog_relative$Name[5]`). Isolated proteins could derive from epiphytic organisms inhabiting the macrophyte surface or from macrophyte tissue underlying them. The contribution of proteins originating from host tissues was evaluated by identifying all the proteins predicted to belong to any taxonomic group within the phyla Streptophyta and Chlorophyta and by calculating their contribution to the number and abundance of all identified proteins. On average, proteins isolated from the surface of *C. nodosa* contained `r filter(metap_cog_streptophyta, cog_abundance=="cog" & host=="Cy")$mean`  ± `r filter(metap_cog_streptophyta, cog_abundance=="cog" & host=="Cy")$sd` \si{\percent} of proteins associated with Streptophyta, contributing to `r filter(metap_cog_streptophyta, cog_abundance=="abundance" & host=="Cy")$mean`  ± `r filter(metap_cog_streptophyta, cog_abundance=="abundance" & host=="Cy")$sd` \si{\percent} of total proteins. Similar to metagenomes, proteins associated with Chlorophyta contributed more to *C. cylindracea* than proteins associated with Streptophyta to *C. nodosa*. Chlorophyta associated proteins comprised `r filter(metap_cog_chlorophyta, cog_abundance=="cog" & host=="Ca")$mean`  ± `r filter(metap_cog_chlorophyta, cog_abundance=="cog" & host=="Ca")$sd` \si{\percent} of all identified proteins in *C. cylindracea*, contributing to `r filter(metap_cog_chlorophyta, cog_abundance=="abundance" & host=="Ca")$mean`  ± `r filter(metap_cog_chlorophyta, cog_abundance=="abundance" & host=="Ca")$sd` \si{\percent} of all protein abundances.

## Discussion
The study of marine macrophyte epiphytic communities using culture independent techniques, such as 16S rRNA analysis, metagenomics and metaproteomics, requires an initial step of biological material isolation. Methods that have been developed for selective isolation of epiphytic biological material can be divided into two groups: (i) procedures involving, prior to extraction, a cell detachment step such as shaking [@Gross2003; @Noges2010], scraping [@Uku2007] and ultrasonication [@Weidner1996; @Cai2014] and (ii) procedures involving a host tissue incubation aiming at direct lysis of epiphytic microbial cells [@Burke2009]. Protocols that include a cell detachment procedure do not provide complete cell removal [@Richter-Heitmann2016; @Cai2014] and in the case of probe ultrasonication tissue disruption can also be observed [@Richter-Heitmann2016]. 
Though, host tissue incubation procedures provide an almost complete cell removal and selective isolation, existing protocols like the one developed for DNA isolation by @Burke2009 use in the extraction buffer a rapid multienzyme cleaner (3M) not available worldwide and without a known composition [@Burke2009]. To circumvent these problems we developed and tested two protocols for selective DNA and protein isolation from marine macrophyte epiphytic communities.

To test if the developed DNA and protein isolation procedures could be applied on a variety of macrophyte species we selected *C. nodosa* and *C. cylindracea* as representatives of seagrass and macroalgal species, on which the procedures were tested. These species especially differ morphologically. While *C. nodosa* leaves are flat, *C. cylindracea* thallus is composed of uneven surfaces [@Kuo2001; @Verlaque2003]. The developed procedures showed an almost complete removal of epiphytic cells from the surfaces of both species comparable to the result of @Burke2009 and indicating that structural differences do not impact the removal efficiency. In addition, isolation protocols were tested in two contrasting season as it is known that macrophytes are harboring more algal epiphytes during autumn and winter [@Reyes2001]. No differences in the removal efficiency was observed between seasons suggesting that protocols can be used on macrophyte samples retrieved throughout the year. Also, no removal differences were observed on samples derived from the same host but from different localities or settlements demonstrating the stability of the protocol in cell removal efficiency.

>>>>

\vspace{20mm}
Marino Korlević^1$*$^, Marsej Markovski^1^, Zihao Zhao^2^, Gerhard J. Herndl^2,3,4^, Mirjana Najdek^1^

@@

4\. Vienna Metabolomics Center, University of Vienna, Austria

@@

\newpage
[]: # (\linenumbers)
\sisetup{mode=text}
\setlength\parindent{24pt}

## Summary
Studies of unculturable microbes often combine methods based on DNA, such as 16S rRNA sequencing and metagenomics, with methods that allow insights into the metabolic status, such as metaproteomics. To apply these techniques to the microbial community inhabiting the surfaces of marine macrophytes it is advisable to perform a selective DNA and protein isolation prior to the analysis to avoid biases due to the host material being present in high quantities. Two protocols for DNA and protein isolation were adapted for selective extractions of DNA and proteins from epiphytic communities inhabiting the surfaces of two marine macrophytes, the seagrass *Cymodocea nodosa* and the macroalga *Caulerpa cylindracea*. Protocols showed an almost complete removal of the epiphytic community regardless of the sampling season, station, settlement or host species. The obtained DNA was suitable for metagenomic and 16S rRNA sequencing, while isolated proteins could be identified by mass spectrometry. The low presence of host DNA and proteins in the samples indicated a high specificity of the protocols. The procedures are based on universally available laboratory chemicals making the protocols widely applicable. Taken together, the adapted protocols ensure an almost complete removal of the macrophyte epiphytic community. The procedures are selective for microbes inhabiting macrophyte surfaces and provide DNA and proteins applicable in 16S rRNA sequencing, metagenomics and metaproteomics.

\newpage
## Introduction
Surfaces of marine macrophytes are colonized by a diverse microbial community whose structure and function are poorly understood [@Egan2013]. As less than 1 % of all prokaryotic species are culturable, molecular methods such as 16S rRNA sequencing, metagenomics and metaproteomics are indispensable to study these organisms [@Amann1995; @Su2012]. Applying these techniques requires an initial isolation step with the purpose of obtaining high quality DNA and proteins.

Biological material (i.e., proteins and DNA) from pelagic microbial communities is usually isolated by collecting cells onto filters and subsequently isolating the target organisms or communities [@Gilbert2009]. If a specific microbial size fraction is aimed sequential filtration is applied [@Massana1997; @Andersson2010]. In contrast, obtaining microorganisms associated to surfaces require either a cell detachment procedure prior to isolation or the host material is co-extracted with the target material. Methods for separating microbial cells from the host include shaking of host tissue [@Gross2003; @Noges2010], scraping of macrophyte surfaces [@Uku2007] or applying ultrasonication [@Weidner1996; @Cai2014]. It was shown that shaking alone is not sufficient to remove microbial cells from surfaces, at least not from plant root surfaces [@Richter-Heitmann2016]. Manual separation methods, such as scraping and brushing are time consuming and subjective, as the detachment efficiency depends on host tissue and the person performing the procedure [@Cai2014]. Ultrasonication was proposed as an alternative method as it is providing better results in terms of detachment efficiency [@Cai2014; @Richter-Heitmann2016]. The downside of this procedure is that complete cell removal is still not obtained and tissue disruption was observed especially after the application of probe ultrasonication [@Richter-Heitmann2016]. An alternative to these cell detachment procedures is the isolation of target epiphytic compounds together with host material [@Staufenberger2008; @Jiang2015]. This procedure can lead to problems in the following processing steps such as mitochondrial and chloroplast 16S rRNA sequence contaminations from the host [@Longford2007; @Staufenberger2008]. In addition, when performing metagenomics and metaproteomics host material can cause biased results towards more abundant host DNA and proteins.

An alternative to these procedures is a direct isolation of the target material by incubating macrophyte tissues in an extraction buffer. After the incubation, the undisrupted host tissue is removed followed by the isolation procedure, omitting host material contaminations. To our knowledge, the only procedure describing a direct and selective epiphytic DNA isolation from the surfaces of marine macrophytes was described by @Burke2009. In contrast to previously described methods, this protocol enables an almost complete removal of the surface community. It was used for 16S rRNA gene clone library construction [@Burke2011a] and metagenome sequencing [@Burke2011]. This method, although providing a selective isolation procedure, uses a rapid multi-enzyme cleaner (3M) that is not available worldwide and the chemical constituents are unknown [@Burke2009]. Also to our knowledge, no selective isolation protocol to perform (meta)proteomics of epiphytic communities associated with marine macrophytes has been developed yet.

In the present study, we adapted a protocol widely used for DNA isolation from filters [@Massana1997] and a protocol used for protein isolation from soils [@Chourey2010; @Hultman2015]. These two adapted methods allowed for a selective extraction of DNA and proteins from epiphytic communities inhabiting the surfaces of two marine macrophytes, the seagrass *Cymodocea nodosa* and the macroalga *Caulerpa cylindracea*. In addition, we tested the removal efficiency of the protocol and the suitability of obtained DNA and proteins for 16S rRNA sequencing, metagenomics and metaproteomics.

@@

To assess the removal efficiency of the DNA and protein isolation procedures, leaves and thalli were examined under a confocal microscope before and after treatments were performed. The modified procedures resulted in an almost complete removal of the surface community of both, *C. nodosa* and *C. cylindracea*. In addition, a similar removal efficiency was observed for communities sampled in contrasting months, December 2017 (\autoref{confocal_december}) and June 2018 (\autoref{confocal_june}). Also, no effect of station, settlement or isolation procedure (DNA or protein) on the removal efficiency was observed (Figs. \ref{confocal_december} and \ref{confocal_june}).

To evaluate whether the obtained DNA is suitable to determine the composition of the microbial community Illumina sequencing of the V4 region of the 16S rRNA was performed. Sequencing yielded a total of `r number_sequences` sequences after quality curation and exclusion of eukaryotic, mitochondrial and no relative sequences. The number of sequences classified as chloroplasts was `r number_chloroplast_sequences`. After excluding these sequences the total number of retrieved reads was `r number_sequences_without_chloroplast`, ranging from `r min_max$min` to `r min_max$max` sequences per sample (\autoref{supp-nseq_notus}). Even when the highest sequencing effort was applied, the rarefaction curves did not level off which is commonly observed in high-throughput 16S rRNA amplicon sequencing (\autoref{supp-rarefaction}). Sequences clustering at a similarity level of 97 \si{\percent} yielded a total of `r no_otus` different OTUs. Taxonomic classification of reads revealed a macrophyte-associated epiphytic community that mainly composed of *Alphaproteobacteria* (`r filter(plot, taxon=="Alphaproteobacteria")$mean` ± `r filter(plot, taxon=="Alphaproteobacteria")$sd` \si{\percent}), *Bacteroidota* (`r filter(plot, taxon=="Bacteroidota")$mean` ± `r filter(plot, taxon=="Bacteroidota")$sd` \si{\percent}), *Gammaproteobacteria* (`r filter(plot, taxon=="Gammaproteobacteria")$mean` ± `r filter(plot, taxon=="Gammaproteobacteria")$sd` \si{\percent}), *Desulfobacterota* (`r filter(plot, taxon=="Desulfobacterota")$mean` ± `r filter(plot, taxon=="Desulfobacterota")$sd` \si{\percent}), *Cyanobacteria* (`r filter(plot, taxon=="Cyanobacteria")$mean` ± `r filter(plot, taxon=="Cyanobacteria")$sd` \si{\percent}) and *Planctomycetota* (`r filter(plot, taxon=="Planctomycetota")$mean` ± `r filter(plot, taxon=="Planctomycetota")$sd` \si{\percent}) (\autoref{community}).

Primers used in this study, as in many other 16S rRNA amplicon procedures, also amplified chloroplast 16S rRNA genes. We observed a high proportion of chloroplast sequences in all analyzed samples (`r filter(plot, taxon=="Chloroplast")$mean` ± `r filter(plot, taxon=="Chloroplast")$sd` \si{\percent}) (\autoref{community}). To determine whether chloroplast sequences originate from the host or eukaryotic epiphytic organisms, we exported SILVA-classified chloroplast sequences and reclassified them using the RDP (Ribosomal Database Project) training set that allows for a more detailed chloroplast classification. The largest proportion of sequences was classified as Bacillariophyta (`r filter(chloroplast, taxon=="Bacillariophyta")$mean` ± `r filter(chloroplast, taxon=="Bacillariophyta")$sd` \si{\percent}) indicating that the DNA removal procedure resulted in only minor co-extracted quantities of host DNA (\autoref{chloroplast}). Chloroplast sequences classified as Streptophyta constituted `r streptophyta_16S$mean` ± `r streptophyta_16S$sd` \si{\percent} of all chloroplast sequences originating from *C. nodosa* samples, while sequences classified as Chlorophyta comprised only `r chlorophyta_16S$mean` ± `r chlorophyta_16S$sd` \si{\percent} of all chloroplast sequences associated with *C. cylindracea* samples.

To determine whether the extracted DNA can be used for metagenomic sequencing, four samples containing epiphytic DNA were selected and shotgun sequenced using an Illumina platform. Metagenomic sequencing yielded between `r format(min(metadata_metagenomic_statistics$raw_R1), big.mark=',')` and `r format(max(metadata_metagenomic_statistics$raw_R1), big.mark=',')` sequence pairs (\autoref{supp-metagenomic_statistics}). Obtained sequences were successfully assembled into contigs whose L50 ranged from `r format(min(metadata_metagenomic_statistics$L50), big.mark=',')` to `r format(max(metadata_metagenomic_statistics$L50), big.mark=',')` bp. In addition, predicted coding sequences were functionally annotated (`r format(min(metadata_metagenomic_statistics$eggnog), big.mark=',')` -- `r format(max(metadata_metagenomic_statistics$eggnog), big.mark=',')` annotated sequences; \autoref{cog}) and taxonomically classified. Functional annotation allowed for an assessment of the relative contribution of each COG (Clusters of Orthologous Groups) functional category to the total number of annotated coding sequences (\autoref{cog}a). Functional categories containing the highest number of sequences were `r cog_relative$"COG Functional Category"[3]` (`r cog_relative$Name[3]`), `r cog_relative$"COG Functional Category"[4]` (`r cog_relative$Name[4]`), `r cog_relative$"COG Functional Category"[5]` (`r cog_relative$Name[5]`), `r cog_relative$"COG Functional Category"[6]` (`r cog_relative$Name[6]`) and `r cog_relative$"COG Functional Category"[7]` (`r cog_relative$Name[7]`). If host DNA is co-extracted with epiphytes it should be detected in large proportions in sequenced metagenomes. However, no large proportions of coding sequences classified as Streptophyta and Chlorophyta were detected (\autoref{supp-metagenomic_taxonomy}). Sequenced metagenomic DNA originating from the surface of *C. nodosa* contained `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCyM" & phylum=="Streptophyta")$cds_relative` \si{\percent} of coding sequences classified as Streptophyta in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCyM" & phylum=="Streptophyta")$cds_relative` \si{\percent} in June 2018. Furthermore, the summed RPKM (Reads Per Kilobase Million) of these sequences constituted `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCyM" & phylum=="Streptophyta")$rpkm_relative` \si{\percent} of total RPKM of all successfully classified sequences in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCyM" & phylum=="Streptophyta")$rpkm_relative` \si{\percent} in June 2018. Similar low proportions of host coding sequences were detected in metagenomic samples originating from the surfaces of *C. cylindracea*. Of all successfully classified coding sequences `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCa" & phylum=="Chlorophyta")$cds_relative` \si{\percent} were classified as Chlorophyta in December 2017 and `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCa" & phylum=="Chlorophyta")$cds_relative` \si{\percent} in June 2018. A relatively higher proportion of RPKM of these sequences than in the case of *C. nodosa* was observed, indicating a higher coextraction of host DNA in *C. cylindracea*. In December, the proportion of RPKM of sequences classified as Chlorophyta was `r filter(metadata_metagenomic_phylum, label=="14/12-17 FCa" & phylum=="Chlorophyta")$rpkm_relative` \si{\percent}, while in June 2018 it reached `r filter(metadata_metagenomic_phylum, label=="19/6-18 FCa" & phylum=="Chlorophyta")$rpkm_relative` \si{\percent}.

To evaluate whether the procedure for protein extraction is suitable for metaproteomic analysis, obtained proteins were trypsin digested and sequenced using a mass spectrometer. Obtained MS/MS spectra were searched against a protein database from sequenced metagenomes. From `r format(min(metap_cog_num), big.mark=',')` to `r format(max(metap_cog_num), big.mark=',')` proteins were identified in isolated protein samples (\autoref{cog}b). In addition, successful identification of proteins allowed for an assessment of the relative contribution of each COG functional category to the total number of identified proteins (\autoref{cog}b). Functional categories containing the highest number of identified proteins were `r metap_cog_relative$"COG Functional Category"[1]` (`r metap_cog_relative$Name[1]`), `r metap_cog_relative$"COG Functional Category"[2]` (`r metap_cog_relative$Name[2]`), `r metap_cog_relative$"COG Functional Category"[3]` (`r metap_cog_relative$Name[3]`), `r metap_cog_relative$"COG Functional Category"[4]` (`r metap_cog_relative$Name[4]`) and `r metap_cog_relative$"COG Functional Category"[5]` (`r metap_cog_relative$Name[5]`). Isolated proteins could originate from epiphytic organisms inhabiting the macrophyte surface and/or from macrophyte tissue underlying them. The contribution of proteins originating from host tissues was evaluated by identifying all the proteins predicted to belong to any taxonomic group within the phyla Streptophyta and Chlorophyta and by calculating their contribution to the number and abundance (NAAF -- Normalized Abundance Area Factor) of all identified proteins. On average, proteins isolated from the surface of *C. nodosa* contained `r filter(metap_cog_streptophyta, cog_abundance=="cog" & host=="Cy")$mean`  ± `r filter(metap_cog_streptophyta, cog_abundance=="cog" & host=="Cy")$sd` \si{\percent} of proteins associated with Streptophyta, contributing to `r filter(metap_cog_streptophyta, cog_abundance=="abundance" & host=="Cy")$mean`  ± `r filter(metap_cog_streptophyta, cog_abundance=="abundance" & host=="Cy")$sd` \si{\percent} of total proteins. Similar to metagenomes, proteins associated with Chlorophyta contributed more to *C. cylindracea* than proteins associated with Streptophyta to *C. nodosa*. Chlorophyta associated proteins comprised `r filter(metap_cog_chlorophyta, cog_abundance=="cog" & host=="Ca")$mean`  ± `r filter(metap_cog_chlorophyta, cog_abundance=="cog" & host=="Ca")$sd` \si{\percent} of all identified proteins in *C. cylindracea*, contributing `r filter(metap_cog_chlorophyta, cog_abundance=="abundance" & host=="Ca")$mean`  ± `r filter(metap_cog_chlorophyta, cog_abundance=="abundance" & host=="Ca")$sd` \si{\percent} to the total protein abundance.

## Discussion
To test whether the developed DNA and protein isolation protocols efficiently detach microbes from the macrophyte surface we selected *C. nodosa* and *C. cylindracea* as representatives of seagrass and macroalgal species. These species differ morphologically. While *C. nodosa* leaves are flat, *C. cylindracea* thalli are characterized by an uneven surface [@Kuo2001; @Verlaque2003]. The developed protocol led to an almost complete removal of epiphytic cells from the surfaces of both species comparable to the result of @Burke2009, indicating that structural differences do not impact the removal efficiency. In addition, isolation protocols were tested in two contrasting seasons, as it is known that macrophytes are harboring more algal epiphytes during autumn and winter [@Reyes2001]. No differences in the removal efficiency was observed between seasons suggesting that these protocols can be used on macrophyte samples retrieved throughout the year. Also, no removal differences were observed on samples derived from the same host but from different locations.

Successful amplification and sequencing of the V4 region of the 16S rRNA gene proved that the isolated DNA can be used to estimate the microbial epiphytic diversity. Taxonomic groups detected in this step can also be often found in epiphytic communities associated with other macrophytes [@Morrissey2019; @Burke2011a]. A problem often encountered in studies focusing on epiphytic communities is the presence of large proportions of chloroplast 16S rRNA sequences in the pool of amplified molecules, especially if the epiphytic DNA was isolated without prior selection [@Staufenberger2008]. These sequences can derive from host chloroplasts or from eukaryotic epiphytic chloroplast DNA. Although the proportion of obtained chloroplast 16S rRNA sequences in our samples was substantial, they derived almost exclusively from eukaryotic epiphytes. High proportion of chloroplast 16S rRNA sequences in studies applying selective procedures that include direct cellular lysis on host surfaces were observed before [@Michelou2013]. It is possible that chloroplast-specific sequences even in these studies originated from eukaryotic epiphytic cells and not from host chloroplasts. Indeed, it is common during 16S rRNA profiling of pelagic microbial communities to observe high proportions of chloroplast sequences [@Korlevic2016; @Gilbert2009]. In addition, a very low proportion of chloroplast 16S rRNA sequences in samples originating from *C. cylindracea* in comparison to *C. nodosa* could be explained by the presence of three introns in the gene for 16S rRNA in some members of the genus *Caulerpa* that could hamper the amplification process [@Lam2016].

High quality DNA is also needed for metagenomics. The obtained number of metagenomic sequences and assembly statistics were comparable to metagenomes and metatranscriptomes derived from similar surface associated communities [@Crump2018; @Cucio2018]. In addition, functional annotation of predicted coding sequences to COG functional categories showed that the obtained metagenomes can be used to determine the metabolic capacity of surface associated communities [@Cucio2018; @Leary2014]. The proportion of coding sequences, including their RPKM, originating from *C. nodosa* metagenomes and classified as Streptophyta was low indicating that the isolation procedure was specific for epiphytic cells. DNA samples isolated from the surface of *C. cylindracea* exhibited a low proportion of Chlorophyta coding sequences, however, their RPKM was higher than in the samples originating from *C. nodosa*. One of the reasons for this elevated RPKM of Chlorophyta sequences in *C. cylindracea* could be the differences in the tissue structure between these two host species. While *C. nodosa* leaves are composed of individual cells, the thallus of *C. cylindracea* is, like in other siphonous algal species, composed of a single large multinucleate cell [@Coneva2015]. The absence of individual cells in *C. cylindracea* could cause a leakage of genetic material into the extraction buffer causing an elevated presence of host sequences in the samples for metagenome analyses.
